Trait' selection from within and between species variations taking in genomics

### 2023-07-17 manuscript/main.tex
Updated manuscript

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heredity}{h^2}
\newcommand{\HereditySpi}{\Heredity_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\widehat{\theta}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{G}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RatePhy}{\sigma^2_{\mathrm{B}}}
\newcommand{\RatePop}{\sigma^2_{\mathrm{W}}}
\newcommand{\RatePopSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRatePop}{\UniDimArray{\RatePop}}
\newcommand{\EstRatePhy}{\widehat{\RatePhy}}
\newcommand{\EstRatePop}{\widehat{\RatePop}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

% Tree
\newcommand{\branch}{\text{b}}
\newcommand{\branchexp}{^{(\branch)}}
\newcommand{\Nbranch}{2 \NbrTaxa - 2}
\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
\newcommand{\Setbranch}{\branch \in \{1, \hdots, \Nbranch \}}
\newcommand{\nodeUp}{\branch^{\uparrow}}
\newcommand{\nodeDown}{\branch^{\downarrow}}
\newcommand{\Ntrait}{K}
\newcommand{\contrast}{\UniDimArray{C}}
\newcommand{\Covariancematrix}{\Sigma}
\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
\newcommand{\Precisionmatrix}{\Omega}
\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
\newcommand{\Identitymatrix}{\BiDimArray{I}}
\newcommand{\brownian}{\mathcal{B}}
\newcommand{\Brownian}{\UniDimArray{\brownian}}
\newcommand{\Scattermatrix}{\BiDimArray{A}}
\newcommand{\Multivariate}{\UniDimArray{Z}}
\newcommand{\indice}{a}
\newcommand{\indiceexp}{^{(\indice)}}
\newcommand{\vecZero}{\UniDimArray{0}}
\newcommand{\vecOne}{\UniDimArray{1}}

@@

\title{A test for diversifying selection for a trait from within and between species variations while integrating for genomic diversity}
\rhead{\scshape Test of diversifying selection}

@@

\keywords{Quantitative genetics \and Phylogenetics \and Population-genetics \and Selection }

@@

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which mode of selection is supported by the observed variation of traits and sequences.
Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.

First, at the micro-evolutionary scale, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
At the species level, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant mode of selection acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
Theoretically, change in traits accumulate linearly with time from divergence for a pair of species, and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
For example, gene expression accumulate linearly with time from divergence, and accumulates faster for trait with large variance~\cite{khaitovich_neutral_2004}.
Such theoretical connection between trait variance and change in mean value can be used to test for a selection of trait using for a pair of species~\cite{walsh_evolution_2018}.
However, to our knowledge, a neutral theoretical expectation for trait changes across several species related by their phylogenetic tree, as a function of within species trait variation has not yet been formulated and tested.

% Phylogenetics (mean trait evolution)
Contrarily, at the macro-evolutionary scale, by accounting for the underlying relationships between species, the mode of selection for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
Indeed, if the trait is neutrally evolving, the change in main trait value along a given branch of the tree should be Normally distributed around the ancestral state (at the beginning of branch), with a standard deviation proportional to the squared root of the divergence time~\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
First, and unfortunately, a better fit for a Brownian process at the phylogenetic scale is not necessarily a proof of the neutral model.
A trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
In such comparative framework, the variation within species is usually discarded, where the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals.
Altogether, methods at the phylogenetic scale easily mis-classify selection and also discards the variance in traits between individuals, leaving out valuable information.

% At the frontier Phylogenetics/Population-genetics
At the frontier between phylogenetics and population genetics, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and sequence variation within and between species.
Importantly, our neutrality index also leverages sequence divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
From a population-geneticist perspective, our study can be seen as the generalization at the macro scale of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
From a phylogeneticist perspective, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} where we set a clear threshold for neutral evolution by leveraging species' sequence polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

@@

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait.
At the individual level, the mutational variance ($\VarMutation = \NbrLoci \Multiply  \E [a^2]$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is also a function of the mutation rate per loci per generation ($\MutationRate$), as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

Altogether, we define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
This ratio allows discarding the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
\begin{align}
    \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heredity$), giving also $\RatePop$ as:
\begin{align}
    \RatePop & = \frac{\Heredity \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.

Thus, after having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, we defined $\RatePhy$ as the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$).
This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePhy$ as:
\begin{align}
    \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
        Within species, genetic variance ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
        Under neutral evolution $\RatePhy$ should equal $\RatePop$.
        Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
    }
    \label{fig:methods}
\end{figure*}

@@

 The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
 At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\RateMut$).
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
Similarly, combined as a ratio, we obtain $\RatePhy$ at the phylogenetic scale, which also depends solely on the genetic architecture of the trait.
Altogether, we thus have:
\begin{gather}
    \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
    \Rightarrow \frac{\RatePhy}{\RatePop} = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
\end{gather}
Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.

@@

\subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}

>>>>

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{G}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

% Tree
\newcommand{\Nbranch}{2 \NbrTaxa - 2}
\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
\newcommand{\Ntrait}{K}
\newcommand{\contrast}{\UniDimArray{C}}
\newcommand{\Covariancematrix}{\Sigma}
\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
\newcommand{\Precisionmatrix}{\Omega}
\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
\newcommand{\Identitymatrix}{\BiDimArray{I}}
\newcommand{\brownian}{\mathcal{B}}
\newcommand{\Brownian}{\UniDimArray{\brownian}}
\newcommand{\Scattermatrix}{\BiDimArray{A}}
\newcommand{\Multivariate}{\UniDimArray{Z}}

@@

\title{Trait' selection from within and between species variations taking in genomics}
\rhead{\scshape Test of trait selection}

@@

\keywords{Quantitative genetics \and Trait evolution \and Selection }

@@

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.

First, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
Gene expression exhibit such accumulation, and divergence in expression accumulates faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for a selection of trait in a pair of species~\cite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Altogether, methods at the phylogenetic scale easily mis-classify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
For population-geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For a phylogeneticist, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

@@

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait, where the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$ as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

Mutations supply new variants, while genetic random drift depletes standing variation.
As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutation segregating divided by the size of the region.
Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

Altogether, we define $\RateWhithin$ as the ratio of genetic variance of the trait ($\VarGenetic$) over nucleotide diversity of any neutral genomic region of interest ($\pi$).
This ratio allows to discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), giving also $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

At generation $\Time$, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.

After having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalised by nucleotide divergence.
        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalised by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the counfunding effect on diversity of mutation rate, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
Similarly, combined as a ratio, $\RateBetween$ depends solely on $\RateMut$
Altogether, we thus have:
\begin{gather}
    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
    \Rightarrow \frac{\RateBetween}{\RateWhithin} = 1.
\end{gather}
Altogether, $\RateBetween / \RateWhithin$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RateBetween$ and $\RateWhithin$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.

@@

\newpage
\subsection{Bayesian estimation}\label{subsec:running-nodetraitsand-readnodetraits}
